---
title: "Introduction to ggplot2"
author: "Lorena Saquicoray"
format: html
editor: visual
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, error = TRUE)

```


```{r}
library(palmerpenguins)
library(tidyverse)
library(skimr)
library(knitr)


data(penguins, package = "palmerpenguins")

glimpse(penguins)
```
## Blank canvas using penguins data. 

```{r}
ggplot(data = penguins
)
```

## New map bill depth to x-axis

```{r}
ggplot(data = penguins, 
       mapping = aes(x = bill_depth_mm))
```

## New map adding bill lenght to y-axis

```{r}
ggplot(data = penguins, 
       mapping = aes(x = bill_depth_mm, y = bill_length_mm))
```

## Scatterplot 2 dimensional with a point layer. 
```{r}
ggplot(data = penguins, 
       mapping = aes(x = bill_depth_mm, y = bill_length_mm))+
geom_point()
```
## Title, subtitle of the plot and, title of the axis. 
```{r}
ggplot(data = penguins, 
       mapping = aes(x = bill_depth_mm, y = bill_length_mm, 
color = species))+
geom_point()+
labs(title ="Penguin bill length by depth", subtitle = "Dimensions for Adelie, Chinstrap, and Gentoo penguins", x = "Bill depth (mm)", y = "Bill length (mm)")
```

We update the title of the legend.
```{r}
ggplot(data = penguins, 
       mapping = aes(x = bill_depth_mm, y = bill_length_mm, 
color = species))+
  
geom_point()+
labs(title ="Penguin bill length by depth", subtitle = "Dimensions for Adelie, Chinstrap, and Gentoo penguins", x = "Bill depth (mm)", y = "Bill length (mm)", color = "Species")
```
## Change the colors 
```{r}
ggplot(data = penguins, 
       mapping = aes(x = bill_depth_mm, y = bill_length_mm, 
color = species))+
geom_point()+
labs(title ="Penguin bill length by depth", subtitle = "Dimensions for Adelie, Chinstrap, and Gentoo penguins", x = "Bill depth (mm)", y = "Bill length (mm)", color = "Species")+ scale_color_viridis_d()
```

## Mapping flipper_length to the alpha asthetic.

```{r}
ggplot(data = penguins, 
       mapping = aes(x = bill_depth_mm, y = bill_length_mm, size = body_mass_g, alpha = flipper_length_mm))+
geom_point()
```
## Plot setting alpha aesthetic to be 0.50
```{r}
ggplot(data = penguins, 
       mapping = aes(x = bill_depth_mm, y = bill_length_mm))+
  geom_point(size =2, alpha = 0.50)

```


```{r}
ggplot(data = penguins)

skim(penguins)

```

## Histogram of penguin body masses 
```{r}
ggplot(data = penguins, 
       mapping = aes(x = body_mass_g))+
  geom_histogram(bins = 30, fill = "grey50", color= "white")+
  labs(title = "Penguin body masses", y = "Frequency", x= "Body mass (g)", caption= "Source: Palmer Station LTER/palmerpenguins package")
```

## box plot of body masses

```{r}
ggplot(data = penguins, 
     mapping = aes(x = body_mass_g, y = species, color = species))+
  geom_boxplot()+
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
    )+

labs(
  title = "Penguin body masses", 
  x = "Body mass (g)", 
  y = "", color ="Species",
  caption = "Source: Palmer Station LTER/palmerpenguins package")
  
```

